### 2011-09-28 papers/smectics/smectics.tex
Added appendix on SPBC
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1456 40876441-3496-0410-a613-a2fca35b2ae2

>>>>

\section{Appendix: sliding periodic boundary conditions}

The simulations of shear flow are undertaken using a mixed finite
difference / lattice Boltzmann approach, where the former is used
to advance the order parameter in time via the Cahn-Hilliard equation,
and the later is used to solve the Navier-Stokes equations.
Coupling between the two is
via the velocity field, which advects the order parameter, and
the thermodynamic pressure tensor (6), the divergence of which
contributes a body force on the fluid. The sliding periodic
boundary conditions of Lees and Edwards\cite{leesedwards} may be
applied to the lattice Boltzmann part of the calculation
\cite{wagnerpagonabarraga,adhikari2005} to generate shear in the system. We
describe here how the sliding periodic boundaries are extended to
include the finite difference part of the calculation.

We consider a finite difference mesh where the order parameter
values are co-located with the density and velocity fields provided
by lattice Boltzmann. Away from the sliding periodic boundaries, the
finite difference discretisation is as one would expect.
In order to ensure conservation of order parameter, as expressed by
\begin{equation}
\partial_t \phi + \partial_\alpha (u_\alpha \phi - M\partial_\alpha \mu) = 0,
\end{equation}
we consider each lattice point to be surrounded by a cubic control volume or
cell
of dimensions $\Delta x = 1$ in two or three dimensions.
We compute convective and diffusive fluxes at the cell boundaries using
linear interpolation of the normal velocity component to the cell face.
For the advective fluxes, the interfacial values of the order parameter
are computed via cubic interpolation weighted in the upwind direction.
This is combined with an Euler forward step for the time integration:
a small Courant number ($u\Delta t/\Delta x$), and hence numerically
stability, is gauranteed by the small Mach number constraint
standard in lattice Boltzmann. The gradient of the chemical potential
is computed at each cell face using values of $\mu$ computed at the
lattice point either side.

A body force on the fluid in the Navier-Stokes equations may be
included in the lattice Boltzmann approach. Again, to ensure conservation
of momentum, the divergence of the thermodynamic presure tensor is
computed based on cell-face values. In practice, this requires a
finite difference stencil
that entends at least $\pm 3$ lattice points in each coordinate direction
(the thermodynamic pressure involves a fourth derivative of the order
parameter).

The sliding periodic boundaries lie (conceptually) along plane
co-incident with the cell faces. Where the finite difference stencil 
overlaps a such a plane, the relative movement between the two sides
of the plane must be accounted for when computing derivatives of the
order parameter, and hence fluxes of order parameter and momentum,
near the planes. This means interpolating values of the order parameter
on each side of the plane to positions corresponding to the lattice
locations on the other side so that the finite difference stencil
can be used normally. For this study, where there are high derivatives
of the order parameter in the free energy, it is important that this
interpolation is at least cubic. Linear interpolation of the order
parameter leads to unphysical oscillations of physical quantities
such as the shear stress related to the magnitude of the velocity jump
across the sliding boundary.
With cubic interpolation, such artefacts are absent. The non-linear
nature of this interpoation means that fluxes of both order parameter
and momentum computed at the plane do not exectly match on each side.
However, this may be corrected at each sliding boundary at each time
step, restoring global order parameter and momentum conservation.
